Deciphering enhancer rules underlying early human neuronal and neural crest development.

### 2025-03-27 manuscript.tex
changes throughout text.

\author[1,2,3,\orcidlink{0000-0001-7907-1247}]{Seppe De Winter}
\author[1,2,3,*,\orcidlink{0000-0002-8006-0315}]{Stein Aerts}

@@

Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols and how they compare to their primary counterparts, at a transcriptomic level. For example, for brain organoids\cite{heIntegratedTranscriptomicCell2024} and endoderm-derived organoids\cite{xuIntegratedTranscriptomicCell2023}. However, such comparisons at the level of the enhancer-code are still limited\cite{zenkSinglecellEpigenomicReconstruction2024, fleckInferringPerturbingCell2023}.\par

@@

In this study, we will determine the enhancer-codes of cell types/states generated by human neural tube organoids, a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021} and compare these codes to those of cell types/states of the head of a four weeks post conception human embryo.

@@

neural crest states (REF). In order to characterize the organoid-derived cell types to a greater level of detail at both

@@

scRNA-seq) at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-110; \textbf{Fig. 1a}),

@@

is evident from the expression of dorso-ventral marker TFs (REF): from ventral to dorsal ARX, NKX2-2, PAX6, IRX3, ZIC1 and
GDF7 (\textbf{Fig. 1d}). In both the organoid and embryo system we obtained multiple early neuronal differentiation states.
A cluster that still expresses the neuronal progenitor marker SOX2, cells that express the early neuronal induction
marker ASCL1 and distinct clusters marked by the expression of GAD1, UNCX and GATA2 (\textbf{Fig. 1e}). Finally,

@@

crest (GRHL2 expression and absence of ZEB2 expression), migratory neural crest (SOX10 and ZEB2 expression), facial
mesenchyme cells (TWIST1 expression) and early differentiation peripheral neurons (NEUROD1 expression). In the embryo,
but not in the organoid, we observed an additional population of cells marked by the expression of SOX10 and PAX2

@@

(SIX3) or anterior spinal cord (HOXB3) identity, we observed a gradient of anterior-to-posterior identities in the
embryo (REF): from anterior to posterior expression of SIX3, EMX2, FOXG1, BARHL1, EN1 and HOXB3 (\textbf{Fig. 1g}).\par

@@

clear that both early differentiation peripheral neurons and to a greater extend early differentiating neurons of the

@@

@@

To gain insight into the sequence characteristics underlying cell type-specific activity of neural tube, neural crest

@@

TF binding motifs as features. For this, we performed motif enrichment analysis on all regions for each topic and

@@

As an illustrative example, we investigated an enhancer located in one of the introns of GLI3 (hs111.0; \textbf{SUPPLEMENTAL

@@

\subsection*{
  Deciphering the enhancer code underlying dorso-ventral patterning of neuronal progenitors
}
Focusing on neuronal progenitors, we identified topics representing cells and regions of different dorso-ventral
identities in both the organoid and the embryo data (\textbf{Fig. 3a}). From ventral to dorsal this is topic 36, 38, 33,
54 and 48 in the organoid data and topic 34, 38, 79, 88 and 58 in the embryo data. As an example we first investigated
two enhancers of \textit{SHH} that are strongly conserved and specific to the floorplate [\textbf{REF}]; SFPE1 and SFPE2.
\textit{SHH} is most strongly expressed in the most ventral topic (33 and 34 resp. in the organoids and the embryo; \textbf{Fig. 3b}).
SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells
belonging to the ventral topic, in both the organoid and embryo data (\textbf{Fig. 3c}). Explaining the accessibility of both
enhancers using DeepNeuralTube reveals two FOX (one partial), one SOX and two RFX (one partial) binding sites in SFPE1
and two FOX and a TEAD binding site in SFPE2 (\textbf{Fig. 3d}).\par

To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we calculated
contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$. We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities (SUPPLEMENTAL FIG). Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code table (\textbf{Fig. 3e}). Both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns
underlying the different dorso-ventral states and the importance of the instances of these patterns is highly correlated across both
models (CORR COEF; \textbf{Fig. 3e}). For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD
and RFX instances is important (\textbf{Fig. 3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA2}, \textit{TEAD1} and \textit{RFX4} in both the organoid and embryo cells (\textbf{Fig. 3g}). In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-2} expression (\textbf{(Fig. 3e- g)}); PAX instances, correlating with \textit{PAX6} expression (\textbf{Fig. 3e-g}) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3e-g}). For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2} expression. Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e}). Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3h}). FOXA2, TEAD1 and RFX4 instances often co-occur in the same genomic region while instances of other patterns seem to be present in separate regions (\textbf{Fig. 3h}).\par

It has been reported that FOXA2 is a pioneer factor. Indeed, when centering on FOXA2 instances in both the organoid and
embryo data we observe a strong footprint when looking at Tn5 cut-sites (\textbf{Fig. 3i} and SUPPLEMENTAL FIG). We then
asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this
TF and regions, bound by FOXA2, that are specific to the floorplate. For this purpose, we downloaded FOXA2 ChIP-seq
peaks in HepG2 and A549 and intersected these with floorplate peaks containing FOXA2 instances. Those floorplate peaks
that overlapped with both ChIP-seq datasets we classified as regions that are generally bound by FOXA2 and those that
did not we classified as floorplate specific. Next, we quantified both the number of FOXA2 instances per region and the
affinity of each individual instance (making use of the FOXA2 motif score calculated using
FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) hypothesizing that the overall affinity of regions that are generally
bound by FOXA2 is higher than floorplate specific regions either because these general regions have more instances or
the affinity of individual instances is higher. Indeed, FOXA2 target regions that are specific to the floorplate mostly
have a single FOXA2 instance while FOXA2 target regions that are general tend to have more instances (\textbf{Fig. 3j}).
Furthermore, in case there is but a single FOXA2 instance in a general FOXA2 target the affinity of this instance tends
to be higher (\textbf{Fig. 3k}). Generating a \textit{de novo} motif from general and specific FOXA2 target regions that
have a single instance, we observe that FOXA2 instances in regions that are general tend to be of the form
\textit{TGTTT\underline{AC}} while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par

@@

Given that we observed cell states from different AP domains in the neuronal progenitors from the embryo (but not
organoid; see \textbf{Fig. 1}) we decided to study the gene regulatory code underlying this patterning further. We
identified seven topics corresponding to different AP domains, from anterior to posterior this is Topic 61, Topic 59,
Topic 31, Topic 62, Topic 70, Topic 52 and Topic 71 (\textbf{Fig. 4a}). We first asked if the AP-program is distinct
from the DV program. In other words, is AP patterning encoded in the same or in different genomic regions as in those in
which DV patterning is encoded? For this purpose, we calculated the pairwise correlation between region-topic
probabilities of all AP and DV topics (\textbf{Fig. 4b}). From this it is clear that both AP and DV patterning as well
as each of their sub-domains are mostly encoded in distinct genomic regions, with the exception of AP Topic 59 and 31
that correlate with the region-topic probability of DV Topic 38 (which corresponds to cells expressing \textit{NKX2-2}; \textbf{Fig. 4b})\par

$DeepNeuralTube_{embryo}$ was able to distinguish the different AP domains but grouped topics of similar domains
together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}). By
explaining the predictions of $DeepNeuralTube_{embryo}$ on the regions of those domains we found four distinct
homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA} corresponding with the expression of respectively \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1} and \textit{HOXB3} [ADD GENE EXPRESSION HEATMAP!]. Finally, SOX2 instances are present in regions of all AP domains (\textbf{Fig. 4e}).\par

Focusing on the two most anterior domains, two similar homeodomain TF binding motifs seem to underlie the differential
accessibility of those domains. Both have the core homeobox motif: \textit{TAAT} but one is flanked by the nucleotides \textit{TAG} while the other is flanked by the nucleotides \textit{CCCT}. We were surprised that this slight difference in the two motifs would be enough for encoding two separate AP domains. To test this hypothesis, we first did a substitution experiment where we replaced all instances of \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} in regions of the most anterior domain and vice versa (\textbf{Fig. 4f}). Indeed, $DeepNeuralTube_{embryo}$ predicts an anterior-to-posterior shift of accessibility of those regions after the \textit{CTAAT\underline{TAG}} to \textit{CTAAT\underline{CCCT}} substitution and vice versa a posterior-to-anterior shift of accessibility after the \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} to  \textit{CTAAT\underline{TAG}} substitution (\textbf{Fig. 4f}). Next, we designed enhancers from scratch by embedding either the  \textit{CTAAT\underline{TAG}} or \textit{CTAAT\underline{CCCT}} motif\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025} and using respectively Topic 61 and Topic 31 as target class (\textbf{Fig. 4g}). [SWITCH PANELS]. Embedding of two instances of either motif was sufficient to reach predicted chromatin accessibility levels of equivalent ranges to the predicted chromatin accessibility levels of genomic regions (\textbf{Fig. 4g}). Next, to eliminate potential model bias we repeated the same experiment but optimized for the opposite class. This is, we optimized for Topic 31 by embedding the \textit{CTAAT\underline{TAG}} motif optimized for Topic 61 by embedding the \textit{CTAAT\underline{CCCT}} motif (\textbf{Fig. 4h}). In this case, the synthetic design failed to converge for the target class but instead regions with high prediction score for the AP domain that corresponds to the motif used for embedding were generated. This is embedding of the \textit{CTAAT\underline{TAG}} motif still generated regions with high prediction score for Topic 61 and embedding of the \textit{CTAAT\underline{CCCT}} motif generated regions with high prediction score for Topic 31 (\textbf{Fig. 4h}).\par

@@

Focusing on neural crest cells, we identified topics representing cells and regions of both pre- and migratory neural
crest as well as facial mesenchyme (\textbf{Fig. 5a}). Similar to the neuronal progenitors, we used
$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each organoid and embryo
topic and generated a code table based on the identified pattern instances (\textbf{Fig. 5b}). Again, both
$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns and the importance of the instances fo
these patterns is highly correlated across both models (CORR COEF; \textbf{Fig. 5b}).\par

Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig. 5b-c}) the importance of TEAD instances correlates with \textit{TEAD4} expression, GRHL with both \textit{GRHL1} and \textit{GRHL2} and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d}). Note that ZEB instances have a positive contribution to the pre-migratory while it has a negative contribution to all other neural crest states (\textbf{Fig. 5b-c}) and moreover all other states of both the neuronal progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig. 5d}). This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par

The migratory neural crest cell state is represented by two organoid topics (Topic 65 and 59) and a single embryo topic (Topic 94). Nuclear receptor, AP2, ZIC (only in organoid topics) and SOX-dimer (organoid Topic 59 but not 65) are important for this state (\textbf{Fig. 5b-c}). We could not link the nuclear receptor instances to a specific TF based on the correlation of expression and the importance score of these instances. The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig. 5d}), ZIC with \textit{ZIC1} to \textit{ZIC5} and the SOX-dimer with expression of \textit{SOX10}. The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothsteinHeterodimerizationTFAP2Pioneer2020}. After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) for which instances of RFX are important (\textbf{Fig. 5b-c}) that correlate with the expression of \textit{RFX4} (\textbf{Fig. 5d}).\par

Finally, the facial mesenchyme cell state is represented by organoid Topic 58 and embryo Topic 91. Instances of a FOX
pattern and a dimer consisting of a homeobox and ebox motif are important for this state (\textbf{Fig. 5b}). The
importance of the FOX instances correlates with the expression of \textit{FOXC2} and that of the dimer motif with the
expression of both \textit{TWIST1} and \textit{ALX4}. The latter is consistent with a recent study where it has been
shown that TWIST1 and ALX4 cooperatively bind this specific dimer motif regulating the expression of genes involved in
facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024}.\par

Next, we investigated which patterns co-occur in the same genomic regions. For this purpose, we calculated the Jaccard
index of genomic regions based on the presence of an instance of each identified pattern (\textbf{Fig. 5e}). Using this
method we found three sets of regions that are predicted to be contain instances of different TFs. One set where
instances of TEAD4, GRHL1/2 and ZEB1/2 co-occur, one where instances of Nuclear receptor, TFAP2A/B/C and ZIC1-5
instances co-occur and one set where instances of FOXC2 and TWIST1/ALX4 co-occur.\par

>>>>

\author[1,2,3,\orcidlink{0000-0001-7907-1247}]{Seppe De Winter}
\author[1,2,3,\orcidlink{0000-0002-1318-2603}]{Camiel Mannens}
\author[1,2,3]{Valerie Christiaens}
\author[1,2,3,*,\orcidlink{0000-0002-8006-0315}]{Stein Aerts}

@@

Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols and how they compare to their primary counterparts, at a transcriptomic level. For example, for brain organoids\cite{heIntegratedTranscriptomicCell2024} and endoderm-derived organoids\cite{xuIntegratedTranscriptomicCell2023}. However, such comparisons at the level of the enhancer-code are still limited\cite{zenkSinglecellEpigenomicReconstruction2024, fleckInferringPerturbingCell2023}. In this study, we will address this gap by determining the enhancer-codes of cell types/states generated by human neural tube organoids, a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021} and compare these codes to those of cell types/states of the head of a four weeks post conception human embryo.\par

@@

@@

neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. In order to characterize the organoid-derived cell types to a greater level of detail at both

@@

scRNA-seq) at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-11; \textbf{Fig. 1a}),

@@

is evident from the expression of dorso-ventral marker TFs (REF): from ventral to dorsal \textit{ARX}, \textit{NKX2-2}, \textit{PAX6}, \textit{IRX3}, \textit{ZIC1} and \textit{GDF7} (\textbf{Fig. 1d}). In both the organoid and embryo system we obtained multiple early neuronal differentiation states.
A cluster that still expresses the neuronal progenitor marker \textit{SOX2}, cells that express the early neuronal induction
marker \textit{ASCL1} and distinct clusters marked by the expression of \textit{GAD1}, \textit{UNCX} and \textit{GATA2} (\textbf{Fig. 1e}). Finally,

@@

crest (\textit{GRHL2} expression and absence of \textit{ZEB2} expression), migratory neural crest (\textit{SOX10} and \textit{ZEB2} expression), facial
  mesenchyme cells (\textit{TWIST1} expression) and early differentiation peripheral neurons (\textit{NEUROD1} expression). In the embryo,
  but not in the organoid, we observed an additional population of cells marked by the expression of \textit{SOX10} and \textit{PAX2}

@@

(\textit{SIX3}) or anterior spinal cord (\textit{HOXB3}) identity, we observed a gradient of anterior-to-posterior identities in the
embryo (REF): from anterior to posterior expression of \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1 }and \textit{HOXB3} (\textbf{Fig. 1g}).\par

@@

clear that both early differentiating peripheral neurons and to a greater extend early differentiating neurons of the

@@

% !!! calculate this corr coefficient for topic based states !!!

@@

To gain insight into the sequence characteristics underlying cell type-specific activity of neural tube-, neural crest-

@@

TF binding motifs as features. For this, we performed motif enrichment analysis on all regions for each topic, to find
known motifs underlying the accessibility of each region. We
also

@@

As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{SUPPLEMENTAL

@@

\subsection*{
  Deciphering the enhancer code underlying dorso-ventral patterning of neuronal progenitors
}
Focusing on neuronal progenitors, we identified topics representing cells and regions of different dorso-ventral
identities in both the organoid and the embryo data (\textbf{Fig. 3a}). From ventral to dorsal this is topic 36, 38, 33,
54 and 48 in the organoid data and topic 34, 38, 79, 88 and 58 in the embryo data. As an example we first investigated
two enhancers of \textit{SHH} that are strongly conserved and specific to the floorplate [\textbf{REF}]; SFPE1 and SFPE2.
\textit{SHH} is most strongly expressed in the most ventral topic (33 and 34 resp. in the organoids and the embryo; \textbf{Fig. 3b}).
SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells
belonging to the ventral topic, in both the organoid and embryo data (\textbf{Fig. 3c}). Explaining the accessibility of both
enhancers using DeepNeuralTube reveals two FOX (one partial) and two RFX (one partial) binding sites in SFPE1
and two FOX, a ZEB and a TEAD binding site in SFPE2 (\textbf{Fig. 3d}).\par

To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we calculated
contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$. We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities (SUPPLEMENTAL FIG). Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3e}). Both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns
underlying the different dorso-ventral states and the importance of the instances of these patterns is highly correlated across both
models (CORR COEF; \textbf{Fig. 3e}). For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD
and RFX instances is important (\textbf{Fig. 3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo cells (\textbf{Fig. 3g} and SUPPLEMENTAL FIG). In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3e- g)}); PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3e-g} and SUPPLEMENTAL FIG) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3e-g} and SUPPLEMENTAL FIG). For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2/3} expression (\textbf{Fig. 3e-g} and SUPPLEMENTAL FIG). Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e} and SUPPLEMENTAL FIG). Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3h}). FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns seem to be present in separate regions (\textbf{Fig. 3h}).\par

It has been reported that FOXA1/2 is a pioneer factor (REF). Indeed, when centering on FOX instances in both the organoid and
embryo data we observe a strong footprint when looking at Tn5 cut-sites (\textbf{Fig. 3i} and SUPPLEMENTAL FIG). We then
asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this
TF and regions, bound by FOXA2, that are specific to the floorplate. For this purpose, we downloaded FOXA2 ChIP-seq
peaks in HepG2 and A549 and intersected these with floorplate peaks containing FOX instances. Those floorplate peaks
that overlapped with both ChIP-seq datasets we classified as regions that are generally bound by FOXA2 and those that
did not we classified as floorplate specific. Next, we quantified both the number of FOX instances per region and the
affinity of each individual instance (making use of the FOX motif score calculated using
FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) hypothesizing that the overall affinity of regions that are generally
bound by FOXA2 is higher than floorplate specific regions either because these general regions have more instances or
the affinity of individual instances is higher. Indeed, FOXA2 target regions that are specific to the floorplate mostly
have a single FOX instance while FOXA2 target regions that are general tend to have more instances (\textbf{Fig. 3j}).
Furthermore, in case there is but a single FOX instance in a general FOXA2 target region the affinity of this instance tends
to be higher (\textbf{Fig. 3k}). Generating a \textit{de novo} motif from general and specific FOXA2 target regions that
have a single instance, we observe that FOX instances in regions that are general tend to be of the form
\textit{TGTTT\underline{AC}} while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par

@@

Given that we observed cell states from different AP domains in the neuronal progenitors from the embryo (but not
organoid; see \textbf{Fig. 1}) we decided to study the gene regulatory code underlying this patterning further. We
identified seven topics corresponding to different AP domains, from anterior to posterior this is Topic 61, Topic 59,
Topic 31, Topic 62, Topic 70, Topic 52 and Topic 71 (\textbf{Fig. 4a}). We first asked if the AP-program is distinct
from the DV program. In other words, is AP patterning encoded in the same or in different genomic regions as DV patterning? For this purpose, we calculated the pairwise correlation between region-topic
probabilities of all AP and DV topics (\textbf{Fig. 4b}). From this it is clear that both AP and DV patterning as well
as each of their sub-domains are mostly encoded in distinct genomic regions, with the exception of AP Topic 59 and 31
that correlate with the region-topic probability of DV Topic 38 (which corresponds to cells expressing \textit{NKX2-1/2}; \textbf{Fig. 4b})\par

$DeepNeuralTube_{embryo}$ was able to distinguish the different AP domains but grouped topics of similar domains
together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}). By
explaining the predictions of $DeepNeuralTube_{embryo}$ on the regions of those domains we found four distinct
homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}. Notably, the patterns corresponding to the two posterior most topics correspond to dimer-motifs. Previously it was already hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through cooperative binding of two TFs\cite{mannChapter3Hox2009}. Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and SUPPLEMENTAL FIG). In contrast, the patterns corresponding to the two anterior most topics correspond to monomer motifs. The presence of instances of these patterns correlates with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior topic (\textbf{Fig. 4d-e} and SUPPLEMENTAL FIG).

Focusing on the two most anterior domains, two similar homeodomain TF binding motifs seem to underlie the differential
accessibility of those domains. Both have the core homeobox motif: \textit{TAAT} but one is flanked by the nucleotides \textit{TAG} while the other is flanked by the nucleotides \textit{CCCT}. We were surprised that this slight difference in the two motifs would be enough for encoding two separate AP domains. To test this hypothesis, we first did a substitution experiment where we replaced all instances of \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} in regions of the most anterior domain and vice versa (\textbf{Fig. 4f}). Indeed, $DeepNeuralTube_{embryo}$ predicts an anterior-to-posterior shift of accessibility of those regions after the \textit{CTAAT\underline{TAG}} to \textit{CTAAT\underline{CCCT}} substitution and vice versa a posterior-to-anterior shift of accessibility after the \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} to  \textit{CTAAT\underline{TAG}} substitution (\textbf{Fig. 4f}). Next, we designed enhancers from scratch by embedding either the  \textit{CTAAT\underline{TAG}} or \textit{CTAAT\underline{CCCT}} motif\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025} and using respectively Topic 61 and Topic 31 as target class (\textbf{Fig. 4g}). Embedding of two instances of either motif was sufficient to reach predicted chromatin accessibility levels of equivalent ranges to the predicted chromatin accessibility levels of genomic regions (\textbf{Fig. 4g}). Next, to eliminate potential model bias we repeated the same experiment but optimized for the opposite class. This is, we optimized for Topic 31 by embedding the \textit{CTAAT\underline{TAG}} motif and we optimized for Topic 61 by embedding the \textit{CTAAT\underline{CCCT}} motif (\textbf{Fig. 4h}). In this case, the synthetic design failed to converge for the target class but instead regions with high prediction score for the AP domain that corresponds to the motif used for embedding were generated. This is embedding of the \textit{CTAAT\underline{TAG}} motif still generated regions with high prediction score for Topic 61 and embedding of the \textit{CTAAT\underline{CCCT}} motif generated regions with high prediction score for Topic 31 (\textbf{Fig. 4h}).\par

@@

Focusing on neural crest cells, we identified topics representing cells and regions of both pre- and migratory neural
crest as well as facial mesenchyme (\textbf{Fig. 5a} and SUPPLEMENTAL FIG). Similar to the neuronal progenitors, we used
$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each organoid and embryo
topic and generated a code-table based on the identified pattern instances (\textbf{Fig. 5b}). Again, both
$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns and the importance of the instances fo
these patterns is highly correlated across both models (CORR COEF; \textbf{Fig. 5b}).\par

Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig. 5b-c}) the importance of TEAD instances correlates with \textit{TEAD3/4} expression, GRHL with \textit{GRHL1/1} and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d} and SUPPLEMENTAL FIG). Note that ZEB instances have a positive contribution to the pre-migratory while it has a negative contribution to all other neural crest states (\textbf{Fig. 5b-c}) and moreover all other states of both the neuronal progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig. 5d} and SUPPLEMENTAL FIG). This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par

The migratory neural crest cell state is represented by two organoid topics (Topic 65 and 59) and a single embryo topic (Topic 94). Nuclear receptor, AP2, ZIC (only in organoid topics) and SOX-dimer (organoid Topic 59 but not 65) are important for this state (\textbf{Fig. 5b-c}). We could not link the nuclear receptor instances to a specific TF based on the correlation of expression and the importance score of these instances. The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig. 5d}), ZIC with \textit{ZIC1} to \textit{ZIC5} and the SOX-dimer with expression of \textit{SOX10} (\textbf{Fig. 5d} and SUPPLEMENTAL FIG). The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothsteinHeterodimerizationTFAP2Pioneer2020}. After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) for which instances of RFX are important (\textbf{Fig. 5b-c}) that correlate with the expression of \textit{RFX3/4} (\textbf{Fig. 5d} and SUPPLEMENTAL FIG).\par

Finally, the facial mesenchyme cell state is represented by organoid Topic 58 and embryo Topic 91. Instances of a FOX
pattern and a dimer consisting of a homeobox and ebox motif are important for this state (\textbf{Fig. 5b}). The
importance of the FOX instances correlates with the expression of \textit{FOXC2/D1/P1/P2} and that of the dimer motif with the
expression of both \textit{TWIST1} and \textit{ALX1/4} and \textit{PRRX1} (\textbf{Fig. 5d} and SUPPLEMENTAL FIG). The latter is consistent with a recent study where it has been
shown that TWIST1 and homeodomain TFs cooperatively bind this specific dimer motif regulating the expression of genes involved in
facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024}.\par

Next, we investigated which patterns co-occur in the same genomic regions. For this purpose, we calculated the Jaccard
index of genomic regions based on the presence of an instance of each identified pattern (\textbf{Fig. 5e}). Using this
method we found three sets of regions that are predicted to contain instances of different set of TFs. One set where
instances of TEAD, GRHL and ZEB co-occur, one where instances of Nuclear receptor, AP2 and ZIC
instances co-occur and one set where instances of FOX and TWIST1/homeodomain co-occur.\par

Recently, a genome wide association study (GWAS) was published identifying 203 single-nucleotide polymorphisms (SNPs)
that are associated with facial variation\cite{whiteInsightsGeneticArchitecture2021}. Given that regions surrounding these SNPs were
enriched for enhancer activity in cranial neural crest cells\cite{whiteInsightsGeneticArchitecture2021} we asked if, using
$DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity. Therefore for all
suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ and calculated the difference in prediction score across both alleles (delta prediction score). At an absolute delta prediction score threshold of 0.5 we identified X SNPs that we deem explainable by either of the two models. Note that only very few (X out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig. 5f}). Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (SUPPLEMENTAL FIG). As an example we show the explanation of rs1555067 which is located upstream of \textit{PRRX1}. The G \textgreater A substitution of this SNP causes a delta prediction score of 0.X for organoid Topic 58 (\textbf{Fig. 5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme cell state. This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp downstream of a partial homeobox site potentially creating a new binding site for TWIST1-homeodomain TF dimers (\textbf{Fig. 5h}).\par
